---
title: "Rooftop solar adoption in Washington DC"
author: "Kumar H"
date: "2024-03-02"
output: 
  html_document:
    code-fold: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# install packages
library(arrow)
library(tidyverse)
library(tidycensus)
library(sf)
library(sandwich)
library(viridis)  # for improved color scales
library(car)
library(lmtest)
library(stargazer)
library(broom)
library(lme4)
library(Matrix)
library(lmtest)
library(recipes)   # For preprocessing steps
library(modeldata) # For data splitting
library(caret)
library(googledrive)
library(kableExtra)


```


As of 2022, solar energy provided about 3.4% of total U.S. utility-scale electricity (U.S. Energy Information Administration, 2024)
Among US cities, Honolulu tops the chart, with per capita solar photovoltaic (PV) capacity installed at 1133 Watts DC/person in 2022. 
Washington DC was at 10th rank by 2022 (leaping up from 17th in 2018.) [(Frontier Group, 2022)](https://frontiergroup.org/resources/shining-cities-2022/)

The most likely reason behind this boost was DC's **Renewable Portfolio Standards (RPS)** policies.
RPS policies are regulatory mandates to increase the production of energy from renewable sources, requiring electric utilities to 
ensure that a specified percentage of the energy they generate or sell comes from renewable sources by a certain date.
Despite the lack of a comprehensive federal RPS, state-level mandates have collectively contributed to a significant increase 
in the country's renewable energy capacity, transforming the energy landscape and paving the way for a more sustainable future.

DC's RPS is one of the country's most ambitious, as underscored by the **Clean Energy DC Omnibus Amendment Act of 2018.**
This act mandates that 100 percent of the District’s electric supply be sourced from renewable generators by 2032. Additionally, 
it sets forth a solar "carve-out" requirement that 5.5 percent of the 2032 electric supply must come from in-District solar, with this percentage increasing
to 10 percent by 2041. This places the District as the third jurisdiction in the U.S., following Hawai'i and California, to
establish a 100 percent renewable energy requirement.

The primary mechanism by which the DC Department of Energy and Environment (DOEE) influences the solar market is through **Solar Renewable Energy Credits (SRECs)**
and **Alternative Compliance Payments (ACP).** 


::: {.callout-tip}
### SRECs and ACP
- **SREC:** Market-based instruments that represent the environmental benefits of generating solar power. 
Each SREC certifies that one megawatt-hour of solar electricity has been generated and can be sold or traded, providing an additional revenue stream for solar energy producers on top of the electricity they sell.

- **ACP:** Penalties that electric utilities or other entities must pay if they fail to meet their renewable energy targets, specifically the portion of those targets that must be met through solar energy. 
:::


There have been several major policy changes in DC that affect solar adoption incentives, as summarized in this RPS report by the DC Public Service Commission (https://dcpsc.org/Orders-and-Regulations/PSC-Reports-to-the-DC-Council/Renewable-Energy-Portfolio-Standard.aspx)
-	**Distributed Generation Amendment Act, 2011**
    -	Amended RPS to increase solar requirements through 2023 to 2.5%.
    - Restricted eligible solar energy resources to those within the District or connected to a feeder serving the District.

- **Renewable Portfolio Standard Expansion Amendment Act, 2016**
    - Increased RPS requirement to 50% by 2032, with solar energy requirement rising to 5%.
    - Modified solar energy RPS compliance fees, setting new rates from 2016 through 2032.

- **CleanEnergy DC Omnibus Amendment Act, 2018**
    - Raised RPS to 100% by 2032, with solar energy requirement increasing to 5.5% by 2032 and 10% by 2041.
    - Extended solar REC banking period from three to five years.

- **Local Solar Expansion Amendment Act, 2022**

    - Increased solar RPS requirement to 15% by 2041.
    - Directed funds from alternative compliance payments to support low-income ratepayer relief and solar policy studies.
    - Adjusted solar energy RPS compliance fees, gradually reducing them from 2024 through 2033, setting a fee of $0.30 per kWh from 2033 through 2041, and $0.10 thereafter.

```{r, include=FALSE}

# inset chatCompletion
# Create the data frame
year <- 2011:2041
tier_one_renewable_percent <- c(4, 5, 6.5, 8, 9.5, 11.5, 13.5, 15.5, 17.5, 20, 26.25, 32.5, 38.75, 45.0, 52.0, 59.0, 66.0, 73.0, 80.0, 87.0, 94.0, 100, 100, 100, 100, 100, 100, 100, 100, 100, 100)
solar_energy <- c(0.4, 0.5, 0.5, 0.6, 0.7, 0.825, 0.98, 1.15, 1.85, 2.175, 2.5, 2.6, 3.0, 3.65, 4.3, 5.0, 5.65, 6.3, 7.0, 7.65, 8.3, 9.0, 9.65, 10.3, 11.0, 11.65, 12.3, 13.0, 13.65, 14.3, 15.0)

data <- data.frame(year, tier_one_renewable_percent, solar_energy)

data$gap <- data$tier_one_renewable_percent-data$solar_energy

# Assuming 'data' is your dataframe
# Add a new column for the act labels
data$Act <- NA  # Initialize with NA
data$Act[data$year == 2011] <- "Distributed Generation Amendment Act of 2011"
data$Act[data$year == 2016] <- "Renewable Portfolio Standard Expansion Amendment Act"
data$Act[data$year == 2022] <- "Clean Energy DC Omnibus Amendment Act"

# Now, plotting with enhancements and adding specific labels
p <- ggplot(data, aes(x = year)) +
  geom_line(aes(y = tier_one_renewable_percent, color = "Tier-One Renewable Percent"), size = 1) +
  geom_area(aes(y = tier_one_renewable_percent), fill = "skyblue", alpha = 0.5) +
  geom_line(aes(y = solar_energy, color = "Solar Energy"), size = 1) +
  geom_area(aes(y = solar_energy), fill = "yellow", alpha = 0.5) +
  geom_point(aes(y = tier_one_renewable_percent), color = "#0099f9", size = 2) +
  geom_label(data = data %>% filter(!is.na(Act)), aes(y = tier_one_renewable_percent, label = Act), nudge_x = 0.25, nudge_y = 0.25, check_overlap = TRUE, size = 3, label.size = 0) +
  scale_color_manual(values = c("Tier-One Renewable Percent" = "#1f77b4", "Solar Energy" = "yellow")) +
  labs(x = "Year", y = "Percent / Energy", title = "Tier-One Renewable Percent and Solar Energy Over Time", color = "Metrics") +
   xlim(2010, 2032) +
  theme_minimal() +
  theme(legend.position = "top",
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10),
        plot.title = element_text(size = 16, face = "bold"),
        axis.title = element_text(size = 14),
        plot.margin = margin(10, 10, 10, 10))

# Print the plot
print(p)


```


### Solar capacity installed has skyrocketed in DC. 
Existing empirical evidence finds that US states that closed off their SREC markets to out-of-state facilites witnessed higher
solar installation, as DC did in 2011. [(Cohen et al 2022)](https://doi.org/10.1111/ajae.12248)

However, as the following animated graph shows, adoption *really* took off around 2016.

```{r, echo=TRUE, eval=FALSE}
#| code-fold: true
#| code-summary: "Sample `gganimate` code"

# Animate the plot
anim <- p + transition_reveal(date) + 
  enter_grow() +  # Animation for entering points
  exit_fade() +   # Animation for exiting points
  ease_aes('cubic-in-out') + # Animation speed
  labs(title = 'Year: {frame_along}', subtitle = "Tier-One Renewable Percent") # Update title with year

# Render the animation
animate(anim, duration = 10, fps = 30, width = 800, height = 600, renderer =gifski_renderer())

```

```{r, echo=FALSE}

# Fetch data for DC
# Convert sf object to GeoJSON
# st_write(joined_df, "solar_acs2019_bg.geojson", driver = "GeoJSON")

library(gganimate)

df_time <- read_csv('data/tot_solar_time.csv')

#create the plot
p <- ggplot(df_time, aes(x = date, y = cum_mw)) +
  geom_line(aes(group = 1), color = "#1f77b4", size = 1.2) +  # Increase line thickness with `size`
  geom_point( color = "#0099f9") + # Draw points
  labs(x = "Year", y = "Capacity added (MW)", title = "Solar Boom in DC") +
   ggthemes::theme_solarized() +
   theme(
    text = element_text(color = "black"), # Make all text black
    plot.title = element_text(size = 26, face = "bold", color = "black"), # Enlarge and bolden the title
    axis.title = element_text(size = 20, face = "bold", color = "black"), # Bolden axis titles
    axis.text = element_text(size = 18) # Adjust size of axis text if needed
  ) 
  # theme(legend.position = "none") # Hide the legend
# Creating the plot with ggplot2

# Animate the plot
anim <- p + transition_reveal(date) + 
  enter_grow() +  # Animation for entering points
  exit_fade() +   # Animation for exiting points
  ease_aes('cubic-in-out') + # Animation speed
  labs(title = 'Year: {frame_along}', subtitle = "Tier-One Renewable Percent") # Update title with year

# Render the animation
animate(anim, duration = 10, fps = 30, width = 600, height = 400, renderer =gifski_renderer())

# Save the animation
# anim_save("dc_solar_anim.gif", animation = anim)


# ts_data_ts <- ts(df_time %>% select(date, cum_mw))
# 
# library(strucchange)
# bp <- breakpoints(ts_data_ts ~ 1)
# 
# plot(bp)
# # Assuming ts_data is your time series data
# cpt.mean <- cpt.mean(ts_data, method="PELT")
# 
#   # Plot the changes
# plot(cpt.mean)
# # Display the plot
# # ggsave("renewable_energy_plot.png", width = 14, height = 7, dpi = 100)

```


Measuring actual generation or consumption, as opposed to capacity installed, is trickier.
In 2022, the total retail electricity sales reported by suppliers in the District were 10.16 million MWh. 
To put this in perspective, Hawaii's consumption in 2022 was around 9 million MWh, despite 
being 60 times larger in area and with 2x the population! ([EIA 2022](https://www.eia.gov/electricity/state/))

"The total number of SRECs retired for compliance was 263,919, which is equivalent to nearly 264,000 MWh and 
represents about 2.6% of the total retail electricity sales for the year"​​. CITE 'DC PSC report'
Of course, SRECs can be held before they are cashed - it indicates that by this date, there has
been *at least* 264,000 MWh generation of solar. 


## The DC Solar Market - Incentives and Costs

In Washington DC, the average solar panel costs [$11,000-$14,000](https://www.solarreviews.com/solar-panels/district-of-columbia) after federal incentives

A household in DC can factor in three types of incentives available for a solar installation :

- Federal rebates

- SREC prices

- Net metering

There is another type of indirect incentive, which might function as a 'stick' to the above 'carrots'. 
This is the [**Building Energy Benchmarking Policy (BEPS)**](https://doee.dc.gov/service/building-energy-performance-standards-beps), which aims to identify and impose financial penalties on large buildings (>10,000 sq feet) with greater than average energy consumption

The following figures describe the outlook for solar generation in the District:

- Technical potential = 1200 MW
- Target by 2040 = 665 MW
- Adoption in 2023 = 201 MW

The above figures are from a report by [Synapse Energy Economics](https://www.synapse-energy.com/sites/default/files/OPCs-Study-on-Future-of-Solar-PV-in-the-District-of-Columbia-18-077.pdf), where it also lays out the factors holding back adoption in DC.
For now, we zoom in on one particular factor laid out in the report:

"Given that **60 percent of households in the District are renter-occupied**, home ownership and upfront costs are likely strong barriers to solar adoption"


## Creating a building-level dataset to examine solar adoption

With the help of DC Open Data’s frequently updated and granular data on buildings, we can merge data solar and building attributes data
for almost all of DC’s >100,000 residential buildings. 

 - [DC PSC Monthly Update of Solar generator certification](https://dcpsc.org/Utility-Information/Electric/RPS/Renewable-Energy-Portfolio-Standard-Program/Monthly-Update-of-Solar-Generator-Certification.aspx_)
   Geocoded dataset of all approved solar panels in Washington DC.

 - [Computer Assisted Mass Appraisal (CAMA)](https://opendata.dc.gov/datasets/c5fb3fbe4c694a59a6eef7bf5f8bc49a)
   
   Building attributes, sourced from tax records properties in Washington, D.C. Contains data on building attributes. 

::: {.callout-important}
## Note: Merging properties with SSL
Linking these two datasets is more complex than a single merge. Solar records are given for an address, but CAMA records are linked to Square Suffix Lot (SSL) numbers, 
which identifies the plot of land on which a building is located. Some lots have multiple buildings, and some buildings/complexes 
span multiple lots.
:::

The following table shows a small sample of the rows and columns in our resulting dataset.

```{r, echo=FALSE}


## AUTHENTICATE Google DRIVE account
# drive_auth()
# 
# 
# # download this to local drive
#                path = "output_data/residential_cross_sec.parquet", overwrite = TRUE)

df <- read_parquet('data/cross_sec_api_12k.parquet')
df$cap_kw <- df$`Capacity (MW)`*1000

df$lat <- df$LATITUDE
df$lon <- df$LONGITUDE
df <- df %>%
  filter(!is.na(lat) & !is.na(lon) 
         & lat != "NA" & lon != "NA") %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
   mutate(cap_kw = replace_na(cap_kw, 0))


# add census block group data
###############

quantile_99 <- quantile(df$cap_kw, 0.99, na.rm = TRUE)


# Preparing subsets
df_with_0 <- df %>% 
  filter(cap_kw == 0) %>%
  select(SSL, STYLE_D, ROOF_D, NBHDNAME, cap_kw, max_sun_hours, min_dc_kwh) %>%
  sample_n(2)

df_without_0 <- df %>%
  filter(`Capacity (MW)` > 0 & `Capacity (MW)` <= quantile_99) %>%
  select(SSL, STYLE_D, ROOF_D, NBHDNAME, cap_kw, max_sun_hours, min_dc_kwh)%>%
  sample_n(3)

# Combining the subsets
df_sample <- bind_rows(df_with_0, df_without_0)

# drop geom
df_sample <- st_set_geometry(df_sample, NULL)

#save
df_sample <- df_sample %>% select(SSL, STYLE_D, ROOF_D, NBHDNAME, cap_kw, max_sun_hours, min_dc_kwh)

# write_csv(df_sample, 'kable_tab.csv')
# 
# # Apply cell_spec to cap_mw for dynamic coloring
df_sample$cap_kw <- cell_spec(df_sample$cap_kw, "html",
                              color = "white",
                              bold = TRUE,
                              background = spec_color(df_sample$cap_kw, end = 0.9, option = "A"))

# Generating the table
knitr::kable(df_sample, "html", escape = FALSE) %>%
  kable_styling(full_width = F, fixed_thead=TRUE,
                bootstrap_options = c("striped", "hover"))


```


The fields `max_sun_hours` is the maximum hours of sunlight the building receives in a year, obtained from the Google Solar API.

## Using Census data to check the home-ownership hypothesis

What does data from the American Community Survey tell us about the characteristics of high-solar adoption areas?
Ideally, we would have joined building-level data of owner vs renter occupation, but in its absence, we can use **block-group** level data
from the ACS 2019 on homeownership rates. 


#### Sample code: `tidyCensus`

```{r, echo=TRUE, eval=FALSE}
#| code-fold: true
#| code-summary: "Sample `tidyCensus` code"

st_acs <- get_acs(variables = c(total_pop = "B02001_001E", 
                                  owner_occupied = "B25003_002E",
                                  total_occupied = "B25003_001E"),
                    geography = "block group",
                    geometry = TRUE,
                    state = "DC", 
                    year = 2019) 
```


```{r, echo=FALSE, message=FALSE, warning=FALSE, results=FALSE}

bg_df <- st_read("data/solar_acs2019_bg.geojson")

# Perform a spatial join: add data from bg_df to df_sample based on their spatial relationship
# zip_data <- zip_data %>%  st_make_valid()
# joined_sf <- st_join(zip_data, bg_df, join = st_contains)

# Improved plot for polygon sf object
# basic_plot <- ggplot(data = bg_df) +
#   geom_sf(aes(fill = total_capacity), color = "black", size = 0.2) +  # Fill polygons and add borders
#   scale_fill_viridis(option = "C", direction = -1) +  # Viridis color scale
#   labs(
#     title = "Per Capita Solar Capacity",
#     fill = "Per Capita kW",
#     caption = "Data Source: ACS 2019, DCPSC 2023"
#   ) +
#   theme_minimal() +
#   theme(
#     legend.position = "right",
#     plot.title = element_text(hjust = 0.5),  # Center the plot title
#     plot.caption = element_text(hjust = 0, size = 8)  # Align caption
#   )
# 
# basic_plot
# 
# model_1 <- lm(pc_capkw ~ own_rate, data = bg_df)
# # Calculate robust standard errors
# # robust_se1 <- vcovHC(model_1, type = "HC1")
# # robust_se1 <- as.vector(coeftest(model_1, vcovHC(model_1, type = "HC1"))[, "Std. Error"])
# -
# Assuming 'joined_sf' is your dataframe with 'own_rate' and 'tot_capkw' variables.

# Create 10 quantiles of ownership rate
bg_df <- bg_df %>%
  mutate(quantile = ntile(own_rate, 20)) %>%
  filter(!is.na(quantile))

# Compute the mean of 'tot_capkw' within each quantile
quantile_avgs <- bg_df %>%
  group_by(quantile) %>%
  summarise(tot_cap = sum(total_capacity),
            tot_pop = sum(total_pop)) %>%
  mutate(cap_kw_pc_1k = 1000*tot_cap/tot_pop)

###########################################################
# Plot the averages as points
ggplot(quantile_avgs, aes(x = as.factor(quantile), y = cap_kw_pc_1k)) +
   # Points with thin black boundary, colored blue
  geom_point(aes(color = as.factor(quantile)), size = 3, shape = 21, fill = "dodgerblue1", color = "black", stroke = 1) +
  # Black and thin line
  geom_smooth(aes(x = as.numeric(as.factor(quantile))), method = "loess", formula = y ~ x, se = TRUE, color = "black", size = 0.8) +
  # Labels and title
  labs(x = "Homeownership Rate (%)", y = "Capacity per 1000 people (KW)", title = "Capacity per 1000 people vs 2019 Ownership rate, block groups") +
  # Minimal theme and angle adjustment for x axis text
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position = "none") +
  # Adjust x-axis labels to show percentage values
  scale_x_discrete(labels = seq(0, 95, by = 5))

# Note: Replace 'joined_sf' with your actual dataframe name.
###########################################################

```


The above graph clearly shows that block groups with a higher share of homes inhabited by owners had higher per capita solar capacity,
expressed as KiloWatts per 1000 people.



## Spatial insights


The fact that the solar data has both the exact location and date of solar panel installation, added with DC Open Data's rich repository
of spatial building-level datasets implies that we can drill down at the hyperlocal level for insights in solar adoption patterns.


```{r, echo=FALSE}
library(leaflet)
library(RColorBrewer)

ward_data <- readRDS("data/ward_solar.rds")
zip_data <- readRDS("data/zipcode_solar.rds")
solar_build <- readRDS("data/solar_build.rds")

# Check the spatial level and generate the map accordingly
pal <- colorNumeric(palette = "YlOrRd", domain = data$total_capacity_mw)
  
# map <- leaflet(zip_data) %>%
#     addProviderTiles(providers$CartoDB.Positron) %>%
#     addPolygons(
#       layerId = ~NAME,
#       fillColor = ~pal(total_capacity_mw),
#       color = "#444444",
#       weight = 1,
#       smoothFactor = 0.5,
#       popup = ~paste("Zip/Ward:", NAME,
#                      "<br>Total Capacity MW:", total_capacity_mw,
#                      "<br>Percentage Solar:", perc_solar),
#       highlightOptions = highlightOptions(weight = 2, 
#                                           color = "white",
#                                           bringToFront = TRUE))

# print(map)

########################################################3
## building

spatialLevel <- "build" # or "zipcode"
# Calculate quantiles for `Capacity (MW)`
solar_build$`Capacity (KW)` <- solar_build$`Capacity (MW)`*1000
quantiles <- quantile(solar_build$`Capacity (KW)`, probs = c(0.01, 0.95))

# Use the quantiles to set the domain for colorNumeric
pal <- colorNumeric(palette = "YlOrRd", domain = c(quantiles[1], quantiles[2]))

# Generate the map with adjusted zoom level and quantile-based color scaling
map <- leaflet(solar_build) %>%
    addProviderTiles(providers$CartoDB.Positron) %>%
    addPolygons(
      layerId = ~Number,
      fillColor = ~pal(`Capacity (KW)`),
      color = "#444444",
      weight = 1,
      smoothFactor = 0.5,
      popup = ~paste("Address:", `Facility Address`,
                     "<br>Total Capacity (KW):", `Capacity (KW)`
                     ),
      highlightOptions = highlightOptions(weight = 2, color = "white", bringToFront = TRUE)
    ) %>%
    setView(lng = -77.04026, lat = 38.93215, zoom = 16) # Adjusted zoom level

# Add a mini-map with the specified position
map %>% addMiniMap(tiles = providers$CartoDB.Positron,
                   toggleDisplay = TRUE,
                   position = "bottomright")

print(map)

```




### Check out this in-progress dashboard here!

<iframe height="400" width="100%" frameborder="yes" src="https://hk939.shinyapps.io/dc_solar_shiny_fnl/"> </iframe>


## Next steps:

Now that the data is consolidated, the next task is to model its diffusion process, and forecast it into the future.

A popular approach is NREL's Distributed Renewable Generation model (`dGen`) model.